Maximum Entropy Spectral Analysis: a case study

### 2021-06-08 paper/mesa.tex
Some edits to the paper- ter
minor rephrasing and a few fixes to the citations

\sschmidt{This could be a footnote?}\textbf{A different approach, which provides a smooth PSD estimation, is to set up a parametric model for the PSD and to fit its parameters to the data with a Markov chain Monte Carlo \cite{Cornish_2015, Littenberg_2015}. Despite being effective, such method is very problem dependent, as it makes heavy assumption on the shape of the PSD. Moreover, it is computationally expensive and it does not come with an handy implementation available to the public. For all those reasons, we do not consider such method in our work.}

@@

Since the order of the AR process is not yet determined by the theory, we opted for an in-depth investigation of several proposals in the literature and found that different loss functions are required for different situations, with the FPE loss function being the most indicated to deal with gravitational wave data. Along these lines, we directly compared the PSDs computed with MESA with the canonical Welch's algorithm. As outlined in Sec.~\ref{sec:Welch_comparison}, MESA provides PSD estimates with smaller variance and better accuracy than Welch algorithm.
The use of MESA is particularly useful for short time series samples, where Welch's method is outperformed in both precision and confidence. \sschmidt{Add here something on accuracy on the PSD estimation, as the referee requires}

This observation suggests a promising avenue to pursue in future developments of gravitational waves data analysis: for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA, the computational cost of MESA is moderate and the inferred PSD is an accurate representation of the true underlying PSD. Hence MESA could be employed for {\it online} PSD estimation during a Bayesian parameter estimation exercise.
%a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalizing over the detector noise PSD.
%\sschmidt{There might be a misunderstanding: I guess the referee is talking about taking into account the PSD errors, while we are talking about dropping the stationariety assumption... The misleading word is `marginalize', I think}
This is possible whenever the log-likelihood of the model for a time series takes the form
\begin{equation}
	\log\mathcal{L}(d_t | \theta) \propto  -\frac{1}{2} \frac{(\tilde{d}(f)- \tilde{x}(f;\theta))^2}{S(f)} - \frac{1}{2}\log\left[\int  df\,S(f)\right]
\end{equation}
where $\tilde{\hphantom{a}}$ denotes the Fourier transform and the signal model $x(t;\theta)$, dependent on some parameter $\theta$, is a prediction for a deterministic signal buried in the observed time series $d(t)$.
In GW data analysis, a typical approach is to estimate the PSD offline on a large batch of data and off-source \cite{lalinference}: this scheme assumes {\it stationary data} on a long timescale (longer than the data under study) and it might not reflect the structure of the noise in the analyzed time slice.
Some alternatives exist for dropping this assumption and modifying the likelihood accordingly \cite{R_ver_2010, R_ver_2011, Edwards_2020, Edwards_2015}.
MESA can add to those an elegant way out: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on the residual $d_t-x_t$.
In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. For this method, we would only need to assume the stationariety of the residual on the batch to analyse\footnote{
Unlike several works in GW data analysis \cite{Littenberg_2013, Biscoveanu_2020, talbot2020gravitationalwave}, this approach does not address the issue of including the PSD uncertainties in the posterior. Instead, it focuses on making minimal assumption about the nature of the data under study.
}, making a lighter assumption on the nature of the data. Furthermore, this method would get, as a bonus, a posterior distribution for the PSD of the analyzed data.
Preliminary studies suggest that this is possible~\cite{martini_thesis}.

>>>>

A alternative approach providing a smooth PSD estimation, is to adopt  a parametric model for the PSD and to fit its parameters to the data with a Markov Chain Monte Carlo \citep[e.g.]{Cornish_2015, Littenberg_2015}. Despite being effective, this method is problem dependent, since it needs to make definite assumptions on the shape of the PSD. Moreover, it can be computationally expensive and it does not come with an handy implementation available to the public. For all the above reasons, we did not consider such method in our work.

@@

Since the order of the AR process is not yet determined by the theory, we opted for an in-depth investigation of several proposals in the literature and found that different loss functions are required for different situations, with the FPE loss function being the most indicated to deal with gravitational wave data. Along these lines, we directly compared the PSDs computed with MESA with the canonical Welch's algorithm. As outlined in Sec.~\ref{sec:Welch_comparison}, MESA provides PSD estimates with smaller variance and better accuracy than Welch algorithm.
The use of MESA is particularly useful for short time series samples, where Welch's method is outperformed in both precision and confidence.

This observation suggests a promising avenue to pursue in future developments of gravitational waves data analysis: for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA, the computational cost of MESA is moderate and the inferred PSD is an accurate representation of the true underlying PSD. Hence MESA could be employed for {\it online} PSD estimation during a Bayesian parameter estimation exercise.
%a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalizing over the detector noise PSD.
%\sschmidt{There might be a misunderstanding: I guess the referee is talking about taking into account the PSD errors, while we are talking about dropping the stationariety assumption... The misleading word is `marginalize', I think}
This is possible whenever the log-likelihood of the model for a time series takes the form
\begin{equation}
	\log\mathcal{L}(d_t | \theta) \propto  -\frac{1}{2} \frac{(\tilde{d}(f)- \tilde{x}(f;\theta))^2}{S(f)} - \frac{1}{2}\log\left[\int  df\,S(f)\right]
\end{equation}
where $\tilde{\hphantom{a}}$ denotes the Fourier transform and the signal model $x(t;\theta)$, dependent on some parameter $\theta$, is a prediction for a deterministic signal buried in the observed time series $d(t)$.
In GW data analysis, a typical approach is to estimate the PSD offline on a large batch of data and off-source \citep{lalinference}: this scheme assumes {\it stationary data} on a long timescale (longer than the data under study) and it might not reflect the structure of the noise in the analyzed time slice.
Some alternatives exist for dropping this assumption and modifying the likelihood accordingly \citep{R_ver_2010, R_ver_2011, Edwards_2020, Edwards_2015}.
MESA can add to those an elegant way out: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on the residual $d_t-x_t$.
In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. For this method, we would only need to assume the stationariety of the residual on the batch to analyse\footnote{
Unlike several works in GW data analysis \citep{Littenberg_2013, Biscoveanu_2020, talbot2020gravitationalwave}, this approach does not address the issue of including the PSD uncertainties in the posterior. Instead, it focuses on making minimal assumption about the nature of the data under study.
}, making a lighter assumption on the nature of the data. Furthermore, this method would get, as a bonus, a posterior distribution for the PSD of the analyzed data.
Preliminary studies suggest that this is possible~\citep{martini_thesis}.
